### 2022-03-01 paper/localpsf.tex
moved old stuff to end. rearranging sections

Once the scattered impulse responses $\impulseresponse_{x_i}$ are computed, we approximate the integral kernel $\Phi(y,x)$ at arbitrary points $(y,x)$ by interpolation. The details of the interpolation will be described in Section SECTION, but the key idea is that
\begin{equation}
\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))
\end{equation}
if $x_i$ is close to $x$. This is shown in Figure FIG. We therefore approximate $\phi_x(y)$ by a weighted linear combination of the values $\phi_{x_i}(y - \mu(x) + \mu(x_i))$ for the $k$-nearest points $x_i$ to $x$. Here $k$ is a user-defined parameter; we use $k=10$. The weights in the linear combination are determined via radial basis function interpolation.

The ability to rapidly evaluate approximate kernel entries $\Phi(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Phi}_H$, of $\Phi$, using the conventional adaptive cross H-matrix construction method. In this method, one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Phi}_H$ into an H-matrix approximation of $\Aop$, denoted $\mathbf{A}_H$. Schematically the overall H-matrix approximation proceeds as follows:
\begin{equation*}
\Aop \longrightarrow V, \mu, \Sigma \longrightarrow \{\impulseresponse_{x_i}\}_{i=1}^r \longrightarrow \boldsymbol{\Phi}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
One may use H-matrix methods to perform further useful linear algebra operations involving $\mathbf{A}_H$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. WHY DO THESE OPERATIONS: KRYLOV SOLVERS, SAMPLING STRATEGIES, CAN'T DO WITH A. These H-matrix methods are fast and scale well to large problems.

@@

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following three approximations, ordered by increasing level of sophistication:
\begin{align}
	\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
	\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
	\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
Approximations \eqref{eq:translate1} and \eqref{eq:translate2} are standard. Approximation \eqref{eq:translate3}, which we use,  is new. These approximations are illustrated in Figure FIG.

Approximation \eqref{eq:translate1} follows from the assumption that $\Phi$ is smooth. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, even if $\Phi$ is smooth (which is typically the case asymptotically) the relevant Lipschitz constant for continuity, and hence the required rank in the low rank approximation, can be large, and this makes algorithms based on assumption \eqref{eq:translate1} highly inefficient. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data. As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank blows up.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

@@

We use approximation \eqref{eq:translate3}, which is an improvement upon \eqref{eq:translate1} and \eqref{eq:translate2} that generalizes both. On one hand, if \eqref{eq:translate1} holds, then $\mu(x) \approx \mu(x_i)$, and so \eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if \eqref{eq:translate2} holds, then $\mu(x_i)-\mu(x) \approx x_i - x$, so again \eqref{eq:translate3} holds. But approximation \eqref{eq:translate3} may continue to hold in situation where neither \eqref{eq:translate1} nor \eqref{eq:translate2} holds. For example, \eqref{eq:translate3} will hold when an operator is locally translation-invariant with respect to a rotated frame of reference, while \eqref{eq:translate2} will not hold in this case (see Figure FIG). Additionally, \eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast, \eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires $y \in \Omega_2$ and $-x+x' \in \Omega_1$ to be added together. We call operators that satisfy \eqref{eq:translate3} ``locally mean-displacement invariant.''

If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,f_{x,y}(y_i)\right), \qquad i=1,\dots,k,
\end{equation}
in the same manner as before, but with $x$ and $y$ now taking opposite roles. We improve the interpolation accuracy at negligible additional cost by combining both of these approximations. Specifically, we combine both approximations by placing function values 
\begin{equation*}
	\left(\widetilde{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widetilde{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of all $2k$ of these points and values at zero.

In detail, define 
\begin{equation*}
\widetilde{x_i} := x_i - x \qquad \text{and} \qquad \widetilde{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
	\widetilde{B} = \begin{bmatrix}
		\widetilde{B}_{xx} & \widetilde{B}_{xy} \\
		\widetilde{B}_{yx} & \widetilde{B}_{xx}
	\end{bmatrix}, \qquad \widetilde{w} = \begin{bmatrix}\widetilde{w}_x \\ \widetilde{w}_y\end{bmatrix}, \qquad \widetilde{f} = \begin{bmatrix}
		\widetilde{f}_x \\ \widetilde{f}_y
	\end{bmatrix},
\end{equation*}
be a $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with the block entries of $\widetilde{B}$ be given by
\begin{align*}
\left(\widetilde{B}_{xx}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{x}_j\|\right), &\qquad
\left(\widetilde{B}_{xy}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{y}_j\|\right), \\
\left(\widetilde{B}_{yx}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{x}_j\|\right), &\qquad \left(\widetilde{B}_{yy}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\Phi(y,x) \approx \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widetilde{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widetilde{y}_i\|\right),
\end{equation}
where $\widetilde{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
	\min_{\widetilde{w}} \frac{1}{2}\|\widetilde{B} \widetilde{w} - \widetilde{f}\|^2 + \frac{\alpha}{2}\|\widetilde{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widetilde{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widetilde{x}_i$ and $\widetilde{y}_j$ in the interpolation such that $\widetilde{x}_i \approx \widetilde{y}_j$, a situation that arises often in practice. Indeed, situations where $\widetilde{x}_i = \widetilde{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widetilde{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard interpolation equation $\widetilde{B} \widetilde{w} = \widetilde{f}$.

@@

@@

>>>>

Once the scattered impulse responses $\impulseresponse_{x_i}$ are computed, we approximate the integral kernel $\Phi(y,x)$ at arbitrary points $(y,x)$ by interpolation. The details of the interpolation will be described in Section SECTION, but the key idea is something that we call \emph{local mean-displacement invariance}, which is the widely applicable assumption that
\begin{equation}
\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))
\end{equation}
if $x_i$ is close to $x$. This is shown in Figure FIG, and discussed in greater detail in Section SEC. We approximate $\phi_x(y)$ by a weighted linear combination of the values $\phi_{x_i}(y - \mu(x) + \mu(x_i))$ for the $k$-nearest points $x_i$ to $x$. Here $k$ is a user-defined parameter; we use $k=10$. The weights in the linear combination are determined via radial basis function interpolation.

The ability to rapidly evaluate approximate kernel entries $\Phi(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Phi}_H$, of $\Phi$, using the conventional adaptive cross H-matrix construction method. In this method, one forms low-rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Phi}_H$ into an H-matrix approximation of $\Aop$, denoted $\mathbf{A}_H$. Schematically the overall H-matrix approximation proceeds as follows:
\begin{equation*}
\Aop \longrightarrow V, \mu, \Sigma \longrightarrow \{\impulseresponse_{x_i}\}_{i=1}^r \longrightarrow \boldsymbol{\Phi}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
One may use H-matrix methods to perform further useful linear algebra operations involving $\mathbf{A}_H$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. WHY DO THESE OPERATIONS: KRYLOV SOLVERS, SAMPLING STRATEGIES, CAN'T DO WITH A. These H-matrix methods are fast and scale well to large problems.

@@

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following three approximations, ordered by increasing level of sophistication:
\begin{align}
	\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
	\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
	\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
Approximations \eqref{eq:translate1} and \eqref{eq:translate2} are standard. Approximation \eqref{eq:translate3}, which we use,  is new. We call operators that satisfy \eqref{eq:translate3} ``locally mean-displacement invariant.'' These approximations are illustrated in Figure FIG.

Approximation \eqref{eq:translate1} follows from the assumption that $\Phi$ is smooth. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, even if $\Phi$ is smooth (which is typically the case asymptotically) the relevant Lipschitz constant for continuity, and hence the required rank in the low rank approximation, can be large, which makes algorithms based on assumption \eqref{eq:translate1} computationally expensive. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data. As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank in the approximation blows up.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

@@

We use approximation \eqref{eq:translate3}, which is an improvement upon \eqref{eq:translate1} and \eqref{eq:translate2} that generalizes both. On one hand, if \eqref{eq:translate1} holds, then $\mu(x) \approx \mu(x_i)$, and so \eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if \eqref{eq:translate2} holds, then $\mu(x_i)-\mu(x) \approx x_i - x$, so again \eqref{eq:translate3} holds. Furthermore, approximation \eqref{eq:translate3} can hold in situations where neither \eqref{eq:translate1} nor \eqref{eq:translate2} holds. For example, \eqref{eq:translate3} will hold when an operator is locally translation-invariant with respect to a rotated frame of reference, while \eqref{eq:translate2} will not hold in this case (see Figure FIG). Additionally, \eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast, \eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires $y \in \Omega_2$ and $-x+x' \in \Omega_1$ to be added together.

@@

\subsubsection{Improvements when $\Aop$ is symmetric}
If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,f_{x,y}(y_i)\right), \qquad i=1,\dots,k,
\end{equation}
in the same manner as before, but with $x$ and $y$ now taking opposite roles. We improve the interpolation accuracy at negligible additional cost by combining both of these approximations. Specifically, we combine both approximations by placing function values 
\begin{equation*}
\left(\widetilde{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widetilde{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of all $2k$ of these points and values at zero.

In detail, define 
\begin{equation*}
\widetilde{x_i} := x_i - x \qquad \text{and} \qquad \widetilde{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
\widetilde{B} = \begin{bmatrix}
\widetilde{B}_{xx} & \widetilde{B}_{xy} \\
\widetilde{B}_{yx} & \widetilde{B}_{xx}
\end{bmatrix}, \qquad \widetilde{w} = \begin{bmatrix}\widetilde{w}_x \\ \widetilde{w}_y\end{bmatrix}, \qquad \widetilde{f} = \begin{bmatrix}
\widetilde{f}_x \\ \widetilde{f}_y
\end{bmatrix},
\end{equation*}
be a $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with the block entries of $\widetilde{B}$ be given by
\begin{align*}
\left(\widetilde{B}_{xx}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{x}_j\|\right), &\qquad
\left(\widetilde{B}_{xy}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{y}_j\|\right), \\
\left(\widetilde{B}_{yx}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{x}_j\|\right), &\qquad \left(\widetilde{B}_{yy}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\Phi(y,x) \approx \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widetilde{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widetilde{y}_i\|\right),
\end{equation}
where $\widetilde{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
\min_{\widetilde{w}} \frac{1}{2}\|\widetilde{B} \widetilde{w} - \widetilde{f}\|^2 + \frac{\alpha}{2}\|\widetilde{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widetilde{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widetilde{x}_i$ and $\widetilde{y}_j$ in the interpolation that are very close to each other, a situation that arises often in practice. Indeed, situations where $\widetilde{x}_i = \widetilde{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widetilde{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard interpolation equation $\widetilde{B} \widetilde{w} = \widetilde{f}$.

@@

\section{OLD CONTENT}
